Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-01 doc/sc2012/soop.tex

There are two major challenges in implementing an efficient task
scheduler for Legion:
\begin{itemize}
\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in the correct order.

\item For performance, Legion must hide the extremely long latencies associated
  with machines that have both distributed memory and many levels of
  memory hierarchy.
\end{itemize}
Our scheduler uses three techniques to achieve both correctness and high performance:
\begin{itemize}
\item Subtasks enjoy an isolation property that allows Legion
to avoid comparing all pairs of tasks for dependences.
Specifically, the only tasks that need to be checked for dependences 
are the immediate subtasks of a common
parent task (see Section~\ref{sec:exec}).

\item Legion uses a {\em deferred execution model} that decouples the issuing
of operations from when operations are performed.  Issued operations wait for other operations on
which they are dependent to complete before executing.  For example, Legion may issue a copy
operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
task $a$ completes and task $b$ will not start until the copy completes.

\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
is distributed and concurrent, and also extracts nested parallelism from subtasks.
\end{itemize}

During execution of a Legion program, each processor runs its own instance of the Legion SOOP.
A processor's SOOP is the Legion runtime system for that processor and handles all requests
for Legion services from tasks executing on the processor.
Legion's SOOP has a five stage pipeline that  separates the management of system resources from actual
task execution.  The five stages are: dependence analysis (Section~\ref{sec:dep}),
distribution (Section~\ref{sec:dist}),
mapping (Section~\ref{sec:map}),
execution (Section~\ref{sec:exec}),
and clean-up (Section~\ref{sec:clean}).
We discuss each in turn.

@@

When a Legion task runs, it may spawn new subtasks to be executed.
Each subtask is {\em registered} with the local SOOP in sequential execution
order, which records the subtask's logical region arguments and associated privileges and coherence properties.  If two subtasks work on indepedent data Legion can schedule
them to out of order, in parallel.  Thus, the first stage in the
execution of a task is dependence analysis between tasks.

Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
$t'$ that may overlap with a logical region used by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
both regions need only read privileges there is never a dependence, but if either
task needs write or reduction privileges, the coherence modes are compared:
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}

An {\tt Dep} entry indicates dependence while {\tt None}
indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
use the same {\em physical region}.  Finally, a {\tt Cont} entry
indicates a dependence unless a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task
waits for the older task to complete before it begins execution.

@@

% *****  isolation, efficient data structure ******

>>>>

There are two major challenges in implementing an efficient task
scheduler for Legion:
\begin{itemize}
\item  For correctness, Legion must guarantee that every pair of {\em dependent} tasks is serialized in program order.

\item For performance, Legion must hide the extremely long latencies associated
  with machines that have both distributed memory and many levels of
  memory hierarchy.
\end{itemize}
Our scheduler uses three techniques to achieve both correctness and high performance:
\begin{itemize}
\item If it were necessary to compare all pairs of tasks for dependences that alone
would be a serial bottleneck in all but the simplest Legion programs.  Fortunately,
subtasks enjoy an isolation property that allows Legion to check only task siblings for dependences
(see Section~\ref{sec:dep}).

\item Legion uses a {\em  software out-of-order processor}, or SOOP, to schedule tasks.  The SOOP 
is pipelined and distributed, and also extracts nested parallelism from subtasks.

\item Legion uses a {\em deferred execution model} that decouples the issuing
of operations from when operations are performed.  An issued operation waits for other operations on
which it is dependent to complete before executing, but a waiting operation does not cause the SOOP
to block, allowing Legion to perform other useful work while the operation is pending.

%  For example, Legion may issue a copy
%operation to move the results of a task $a$ to the place where a task $b$ will take the copied data as
%an argument.  Tasks $a$ and $b$ and the copy can all be issued, but the copy will not start until
%task $a$ completes and task $b$ will not start until the copy completes.

\end{itemize}

Each processor in the machine runs an instance of the SOOP, which handles all requests
for Legion services from tasks executing on the processor.  The SOOP is analagous to the out-of-order
instruction scheduler in superscalar processors, but working at the granularity of tasks and their region
arguments instead of instructions and their register arguments.
Legion's SOOP has a five stage pipeline: dependence analysis (Section~\ref{sec:dep}),
distribution (Section~\ref{sec:dist}),
mapping (Section~\ref{sec:map}),
execution (Section~\ref{sec:exec}),
and clean-up (Section~\ref{sec:clean}).
We discuss each in turn.

@@

When a Legion task runs, it may spawn new subtasks to be executed.  We refer to the
subtasks as {\em children} of the {\em parent} task.  The children of a parent task
are {\em siblings} of each other.

When a parent task spawns a child task, the child is {\em registered} with the SOOP 
on which the parent is executing; registration records the subtask's logical region
arguments and associated privileges and coherence properties.
Children are registered in the sequential execution order in which the parent task
spawns them.

If two children have a dependence between their region arguments, the
SOOP will schedule them to run serially to preserve the program's
sequential semantics.  If two child tasks have no dependence then
Legion can schedule them to execute out of order, in parallel.  Thus,
the first stage in the execution of a task is dependence analysis
between tasks.

\subsubsection{Dependences}
Conceptually, detecting dependences between a newly registered task $t$ and a previous task 
$t'$ requires comparing the two sets of logical regions accessed.  For each logical region used by
$t'$ that may overlap with a logical region used by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
both regions need only read privileges there is never a dependence, but if either
task needs write or reduction privileges, the coherence modes are compared:
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}

A {\tt Dep} entry indicates dependence while {\tt None}
indicates independence.  A {\tt Same} entry is a dependence unless the two tasks
use the same physical instance of the logical region.\footnote{For example, if two tasks wish to acess the region with atomic coherence, then if they are working on the same physical copy of the data atomicity can be guaranteed using, e.g., locking.  If they are working on different copies of the data then the only safe execution strategy for Legion is to serialize the two tasks.}
A {\tt Cont} entry
indicates a dependence unless a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task
waits for the older task to complete before it begins execution.

@@

\subsubsection{Isolation}
The reader may wonder why we only compare sibling tasks for
dependences: can't arbitrary pairs of tasks have dependences?  The
answer is ``yes'', but the following lemma shows that
any non-sibling tasks are always detected first as a dependence between
siblings that are registered earlier in execution!
\begin{lemma}
\label{lem:isolation}
\rm
Let $t_1$ and $t_2$ be two distinct, non-sibling tasks other than the root task with parents $p_1$ and $p_2$, respectively.
If $t_1$ and $t_2$ have a dependence, then $p_1$ and $p_2$ have a dependence.
\end{lemma}
\begin{proof}
Recall from Section~\ref{sec:ex} that the parents' region arguments include the children's region arguments and have
only more permissions.
\end{proof}
Lemma~\ref{lem:isolation} justifies checking dependences only between sibling tasks, because any dependences between 
arbitrary tasks are detected first between children of the tasks' least common ancestor.  Legion
guarantees that all dependence analysis is performed for all children of a task $t$ on the instance of the SOOP where $t$
runs, which by Lemma~\ref{lem:isolation} is sufficient to detect all required dependences across all tasks.

\subsubsection{Efficient Dependence Analysis}
